---
title: "Which Penguin Species Has the Biggest Bill?"
subtitle: "Built with: Python (pandas), R (ggplot2), Quarto, reticulate"
author: "Ahmed Muhaimin Choudhury"
date: 2026-09-25
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/), Palmer Station Antarctica LTER, released under CC0.

This post uses **both Python and R in one document** via the `reticulate` package. Python (pandas) handles the data summary; R (ggplot2) handles the visualization. A pandas DataFrame is passed from Python into R via `py$`. The analysis tries to find out which penguin species has the biggest bill in the palmer penguin dataset.

## Checking Environment

```{r}
#| label: env-check-r
#| message: false
#| warning: false
library(reticulate)
use_virtualenv(here::here(".venv"))  # point reticulate at the project venv

R.version.string
.libPaths()[1]
```

```{python}
#| label: env-check-python
import sys
print(sys.executable)  # confirm reticulate is using the project venv
```

Both outputs above should point inside this repository's `renv/` and `.venv/` — confirming the two language environments are project-local.

## Data summary (Python)

```{python}
#| label: py-summary
#| message: false
#| warning: false
from palmerpenguins import load_penguins
import pandas as pd

# drop rows missing species or bill measurements
penguins = load_penguins().dropna(subset=["species", "bill_length_mm", "bill_depth_mm"])


# mean bill length + depth per species
summary_df = (
    penguins
    .groupby("species")[["bill_length_mm", "bill_depth_mm"]]
    .mean()
    .round(2)
    .reset_index()
)
```

## Handoff and table in (R)

The `summary_df` computed in Python is now available in R as `py$summary_df`. It arrives as a native R data frame. We use this to render the table using knitr::kable, so that the table looks prettier than reticulate's basic looking data dump.

```{r}
#| label: tbl-py-summary
#| tbl-cap: "Mean bill length and depth across penguin species."
knitr::kable(py$summary_df) #knitting a pretty looking table
```
Adelie penguins have the shortest bills. Chinstraps and Gentoos have similar bill lengths on average, but Gentoos have noticeably shorter bill depth.


## Visualization in R
Now, we will be using R ggplot2 to visualize the data and take a different look at it. 


```{r}
#| label: fig-bill-dimensions
#| fig-cap: "Mean bill length and depth across penguin species. Data aggregated in Python (pandas), visualized in R (ggplot2)."
#| fig-width: 7
#| fig-height: 4.5
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(ggplot2)

# reshape the pandas DataFrame from Python into long format for grouped bars
plot_data <- py$summary_df |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm),
    names_to = "measurement",
    values_to = "mm"
  ) |>
  mutate(
    measurement = recode(measurement,
      bill_length_mm = "Bill length",
      bill_depth_mm = "Bill depth"
    )
  )

ggplot(plot_data, aes(x = species, y = mm, fill = measurement)) +
  geom_col(position = "dodge", width = 0.7) +
  labs(
    x = "Species",
    y = "Mean value (mm)",
    fill = NULL
  ) +
  theme_minimal()
```

## Inference

The plot shows why bill length alone would misclassify Gentoos. Chinstraps and Gentoos both have long bills (~48mm), and Gentoos have distinctly shorter, blockier bills. But a depth-to-length ratio that sets them apart. Bill length identifies Adelie easily; separating Chinstrap from Gentoo needs depth too. This is a small but concrete example of how interoperability strengthens our modelling by bridging the best of features from 2 different languages. 